Physcraper: A Python package for continually updated phylogenetic trees using the Open Tree of Life

### 2021-04-29 sources/bmc_template_submission/bmc_article.tex
tiny edits

Data such as
geographical location, fossil ranges, and genetic and phenotypic information increasingly
available in public biological databases constitute an amazing
resource for scientific discovery \cite{baxevanis2015importance}.
One of the main challenges for automatic
integration of data across biological databases are varying taxonomic idiosyncrasies.
To address this challenge, the Open Tree of Life project (OpenTree)
created a unified taxonomy for automatic taxonomic name standardization, by integrating taxonomic
data from several databases \cite{rees2017automated}, including the USA National
Center for Biodiversity Information (NCBI) taxonomy \cite{federhen_ncbi_2012, schoch_ncbi_2020},
and the Global Biodiversity Information Facility (GBIF) \cite{gbif_secretariat_gbif_2019}, among many others.
All of OpenTree tools and methods are carried out via OpenTree's Application Programming Interfaces (APIs),
that are open as services for use by the general public \cite{opentreeAPIs}.
Physcraper leverages on existing OpenTree's taxonomic APIs to
standardize taxon names in any phylogeny, providing
a framework for automatic connection of updated phylogenies with data from any biological database.

@@

%% Biological plausibility of example??
Our example application of Physcraper to update a phylogeny of the genus \textit{Ilex} is based on a
single marker, so we expect for it to be not as well resolved as phylogenies resulting from analyses
that used multiple markers. Although not perfect,
we think the Physcraper updated \textit{Ilex} phylogeny seems biologically reasonable
in different ways. All samples
corresponding to the ingroup cluster together forming a monophyletic group (Figure 2A), and
samples belonging to the same \textit{Ilex} species also form
monophyletic groups (Figure 2B). A notable exception is samples of the species \textit{Ilex theeizans}, which appear as
non-monophyletic in the updated phylogeny as well as in the original Gottlieb2005 phylogeny.
%This suggests that the samples might represent two different species, but further
analyses should be conducted
A visual comparison of the Yao2020 phylogeny and the original Gottlieb2005 phylogeny suggests
that the relationships within the genus \textit{Ilex} are still being actively
determined, and that increased taxon sampling might be key to resolve them.

>>>>

Data such as
geographical location, fossil ranges, and genetic and phenotypic information increasingly
available in public biological databases constitute an amazing
resource for scientific discovery \cite{baxevanis2015importance}.
One of the main challenges for automatic
integration of data across biological databases are varying taxonomic idiosyncrasies.
To address this challenge, the Open Tree of Life project (OpenTree)
created a unified taxonomy for automatic taxonomic name standardization, by integrating taxonomic
data from several databases \cite{rees2017automated}, including the USA National
Center for Biodiversity Information (NCBI) taxonomy \cite{federhen_ncbi_2012, schoch_ncbi_2020},
and the Global Biodiversity Information Facility (GBIF) \cite{gbif_secretariat_gbif_2019}, among many others.
All of OpenTree tools and methods are implemented via OpenTree's Application Programming Interfaces (APIs),
as open programmatic access services for use by the general public \cite{opentreeAPIs}.
Physcraper leverages on existing OpenTree's taxonomic APIs to
standardize taxon names in any phylogeny, providing
a framework for automatic connection of updated phylogenies with data from any biological database.

@@

%% Biological plausibility of example??
Our case study application of Physcraper to update a phylogeny of the genus \textit{Ilex} is based on a
single marker, so we expect for it to be not as well resolved as phylogenies resulting from analyses
that used multiple markers. Although not perfect,
we think the Physcraper updated \textit{Ilex} phylogeny seems biologically reasonable
in different ways. All samples
corresponding to the ingroup cluster together forming a monophyletic group (Figure 2A), and
samples belonging to the same \textit{Ilex} species also form
monophyletic groups (Figure 2B). A notable exception is samples of the species \textit{Ilex theeizans}, which appear as
non-monophyletic in the updated phylogeny as well as in the original Gottlieb2005 phylogeny.
% This suggests that the samples might represent two different species, but further
% analyses should be conducted.
A visual comparison of the Yao2020 phylogeny and the original Gottlieb2005 phylogeny suggests
that the relationships within the genus \textit{Ilex} are still being actively
determined, and that increased taxon sampling might be key to resolve them.
